\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{10.1145/1122445.1122456}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural
  Gaze Detection}{June 03--05, 2018}{Woodstock, NY}
\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
  June 03--05, 2018, Woodstock, NY}
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/18/06}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\usepackage{geometry}
%\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}

% Basics
%\usepackage{fixltx2e}
%\usepackage{url}
%\usepackage{fancyvrb}
%\usepackage{mdwlist}  % Miscellaneous list-related commands
\usepackage{xspace}   % Smart spacing
%\usepackage{supertabular}


% https://www.nesono.com/?q=book/export/html/347
% Package for inserting TODO statements in nice colorful boxes - so that you
% won’t forget to ﬁx/remove them. To add a todo statement, use something like
% \todo{Find better wording here}.
%\usepackage{todonotes}

%for backtick `
\usepackage{textcomp}

%% Math
%\usepackage{amsmath}
\usepackage{amsthm}
%\usepackage{amssymb}
%\usepackage{bm}       % Bold symbols in maths mode

%% Theoretical computer science
\usepackage{stmaryrd}
\usepackage{mathtools}  % For "::=" ( \Coloneqq )

%% Code listings
\usepackage{listings}
\usepackage{xcolor}

%for cref -- celever referencing
\usepackage{cleveref}

\usepackage{amsbsy}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    %backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=1pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2,
    %xleftmargin=.2\textwidth, xrightmargin=.2\textwidth,
    numbers=none,
    language=Java
}
\lstset{style=mystyle}

%% Font
\usepackage[euler-digits,euler-hat-accent]{eulervm}


\usepackage{ottalt}

\renewcommand\ottaltinferrule[4]{
  \inferrule*[right={#1}]
    {#3}
    {#4}
}

\newcommand\mynote[3]{\textcolor{#2}{#1: #3}}
\newcommand\bruno[1]{\mynote{Bruno}{red}{#1}}
\newcommand\baber[1]{\mynote{Baber}{blue}{#1}}
\newcommand\snow[1]{\mynote{Snow}{orange!80!black}{#1}}
\newcommand\ningning[1]{\mynote{NN}{purple}{#1}}

\newcommand{\name}{$\lambda_{u}$\xspace}
%\newcommand{\dut}{\textsf{Disjoint Union Types}\xspace}
\newcommand{\cal}{$\lambda_{u}$\xspace}
\newcommand{\typeof}{$typeof$\xspace}
\newcommand{\Typeof}{$Typeof$\xspace}

% Ott includes
\inputott{ott-rules}

\title{Union Types with Disjoint Switches}

%set bibliography
\setcitestyle{authoryear,round}
\bibliographystyle{plainnat}% the mandatory bibstyle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \author{Xuejing Huang}
%% \orcid{0000-0002-8496-491X}
%% \affiliation{
%%   \institution{The University of Hong Kong}
%% }
%% \email{xjhuang@cs.hku.hk}


\author{Bruno C. d. S. Oliveira}
%% \authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%% \orcid{}             %% \orcid is optional
\affiliation{
  %% \position{Position1}
  %% \department{Department1}              %% \department is recommended
  \institution{The University of Hong Kong}            %% \institution is required
  %% \streetaddress{Street1 Address1}
  %% \city{City1}
  %% \state{State1}
  %% \postcode{Post-Code1}
  %% \country{Country1}                    %% \country is recommended
}
\email{bruno@cs.hku.hk}          %% \email is recommended
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{abstract}
Union types allow expressing terms with alternative types. Several forms of
union types have been investigated in the research literature, and
union types are nowadays a common feature in many modern programming
languages.
%The Ceylon programming language is interesting in that it
%supports union types, with an innovative elimination construct that has
%not been previously studied in the rese
This paper investigates a particular formulation of union types,
inpired by the Ceylon programming language, with an elimination
construct that enables case analysis (or switches) on types.  The
interesting aspect of such construct is that each clause must operate
on \emph{disjoint} types. By using such elimination construct, it is
possible to ensure that all possible cases are
handled and that none of the cases overlap. In turn, this means that
the order of the cases does not matter and reordering the cases has no
impact on the semantics, which can aid program undertanding and
refactoring. While implemented in the Ceylon language, such construct
with disjoint switches has not been formally studied in the research
literature, although a related notion of disjointness has been studied
in the context of \emph{disjoint intersection types}.

We study union types with disjoint switches formally and in a language
independent way.  We present the \emph{union calculus} (\cal) that
includes disjoint switches and prove several results, including type
soundness and determinism. The notion of disjointness in \cal is
interesting in that it is in essence the dual notion of disjointness
for intersection types.  However, there are challenges that arize for
disjointness when combining union and intersection types.  We also
study several extensions, including intersection types and
distributive subtyping and show that such extensions retain all the
desired properties. All the results about \cal and its extensions
are formalized in the Coq theorem prover.
\end{abstract}

\begin{comment}
    With advance types
such as intersection types and union types, it has become a challenge
to define a robust, type-safe, coherent and deterministic type
system. One often has to compromise on one property to attain
another. Union types with pattern matching on types add significant
expressive power in programming language. Function overloading can
simply be expressed in a single function with the help of union types
and case analysis on types. Intersection types incorporate many
interesting and advance features that are not easy to implement in
classical OOP model.  This study proposes a novel calculus with all
aforementioned properties for pattern matching with union types and
intersection types. The calculus proposed in this study is named as
\cal.  Outline idea in \cal is to allow only non-overlapping or
disjoint types in case expressions.
\end{comment}

\maketitle

%\ottdefnsOrdinary

%\ottdefnsBottomLike

%\ottdefnsDisjointness

%\ottdefnsSubtyping

%\ottdefnsTyping

%\ottdefnsReduction

\input{sections/introduction.mng}
\input{sections/overview.mng}
\input{sections/union.mng}
\input{sections/intersection.mng}
\input{sections/discussion.mng}
\input{sections/related.mng}
\input{sections/conclusion.mng}

\bibliography{paper}


\end{document}
